Bootstrap learning via modular concept discovery

### 2013-04-22 drafts/ijcai2013/ijcai13.tex
Got ijcai13.tex to compile.

\usepackage{ijcai13}
\usepackage{times}
\usepackage{amsmath} 
\usepackage{latexsym} 
\usepackage{ dsfont }
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{tikz}
\usepackage{tikz-qtree}
\usepackage{float}
\usepackage{circuitikz}
\usepackage{graphicx}
\usepackage[font=small]{subcaption}
\usepackage[font=small]{caption}
\usepackage{alltt}
\usepackage{dashrule}
\usepackage{array}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usetikzlibrary{arrows,automata, positioning, patterns,backgrounds}
\usetikzlibrary{arrows,shapes.gates.logic.US,shapes.gates.logic.IEC,calc}

@@

\begin{figure}
%%  \centering
  \begin{subfigure}[Before]{0.45\linewidth}
    \begin{tikzpicture}[scale=.8,
        transform shape, level/.style={sibling distance=5mm/#1, font=\scriptsize}
        ]
      \node [font=\scriptsize] (top) {$\mathbb{R}$}
      child {node [above=.3cm, left=.1cm, text width=1.8cm] (a) {$(\mathbb{R} \rightarrow \mathbb{R}) \rightarrow \mathbb{R} 
          \rightarrow \mathbb{R}$}
        child {node [left=.5cm, below=1cm, text width=3.0cm] 
          {$ \mathbf{S} : (t_2 \rightarrow t_1 \rightarrow t_0)  \rightarrow (t_2 \rightarrow t_1) 
            \rightarrow \\ t_2 \rightarrow t_0$}}
        child {node [right=0cm, text width=2cm] {$ \mathbf{*}: \mathbb{R} \rightarrow \mathbb{R} \rightarrow \mathbb{R}$}}
      }
      child {node [below right=-.5cm and .2cm] (b) {$\mathbb{R} \rightarrow \mathbb{R}$}
        child {node [above=.5cm] {$ \mathbf{I}: t_0 \rightarrow t_0 \rightarrow t_0$}}
      }
    \end{tikzpicture}
    \caption{    \label{fig:clbintree}}
  \end{subfigure}
  \begin{subfigure}[After]{0.5\linewidth}
    \begin{tikzpicture}[scale=.7, transform shape, 
        level/.style={sibling distance=10mm/#1},
        every edge/.style={draw, thin}
      ]
      \node [circle,draw] (a) {}
      child {node [left=0cm] (b) {$f_1$}}
      child {node [left=.2cm] (c) {\dots} edge from parent[draw=none] }
      child {node [left=.4cm] (e) {$f_n$}}
      child {node [circle,draw, left=1cm] (d) {} edge from parent[very thick]
        child {node [circle, draw, left=.25cm] (l) {} edge from parent[draw=none]
          child {node [left=.5cm] (b) {$f_1$}} 
          child {node [left=.3cm] (c) {\dots} edge from parent[draw=none] }
          child {node [left=.3cm] (e) {$f_n$} edge from parent[very thick]}
          child {node [circle, draw, left=.25cm] (and2) {}
            child {node [circle, draw, left=.25cm](l2) {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
            child {node [circle, draw] (r2)  {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
          }
        }
        child {node [circle, draw, right=.25cm] (r) {} edge from parent[draw=none]
          child {node [circle, draw, right=.25cm] (and3) {}
            child {node [circle, draw ](l3) {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
            child {node [circle, draw, right=.2cm ] (r3)  {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
          }
          child {node [right=.3cm] (b) {$f_1$} edge from parent[very thick]} 
          child {node [right=.4cm] (c) {\dots} edge from parent[draw=none] }
          child {node [right=.6cm] (e) {$f_n$}}
        }
      };

@@

\end{subfigure}
%\vspace{-.3cm}
\caption{Learning curves from Boolean function learning
  experiments. (a) Learning curves for various frontier sizes on task
  set of circuit based Boolean functions. (b) 10
  learning curves for frontier size of 2000 on a task set consisting
  of all Boolean functions of cardinality 3 or smaller. Note how most
  of the trajectories get stuck in a local minimum around
  $50\%$.\label{fig:boolLearningCurves}

}
%\vspace{-.5cm}
\end{figure}

@@

@@

At their best, human processes of discovery clearly go beyond
what our algorithm is capable of.  The algorithm can get stuck in
plateaus after a rapid initial increase in performance, due to its
circular nature: it only inspects the region of the solution space
that is typical of solutions already found.  Humans can become stuck
in similar ways, but at least sometimes they realize that they are
stuck and attempt to find another class of concepts for the remaining
unsolved tasks.  Our algorithm requires, as human learners do, a
sufficiently graded ``curriculum'', or spectrum of problems to solve --
simple problems provide necessary stepping stones to building complex
representations.  Also like humans, our algorithm can adaptively
``self-pace'' its ways through the curriculum, figuring out which
problems it is ready to tackle when, which problems are appropriately
difficult given what it currently knows.  Yet people can sometimes
more intelligently compose and modify the set of problems under
consideration, designing their own stepping stones to expand the set
of problems they can solve.  Making bootstrap learning systems smarter
in these more human-like ways is a prime target for future research.
%% \subsection{Digital arithmetic}
%% Having demonstrated the behavior of the E-C algorithm on symbolic
%% regression, we analyze its performance on tasks involving digital
%% arithmetic, a class of problems with known modular structure. The
%% general goal is to discover boolean circuits consisting of a limited
%% set of gates that perform arithmetic operations on binary number
%% representations.

%% \section{Discussion}
%% This work attempts to bring together ideas from various areas of
%% research -- hierarchical Bayes, learning to learn, minimum description
%% length learning -- to provide a proof of concept that learning
%% abstractions from programs via compression can rapidly improve search
%% performance. We emphasize this by focusing on domains in which the
%% solution space is not smooth and the error signal is all or
%% none. Thus, the main source of information for learning is the modular
%% and compositional structure of the solution space.

%% Of course, the representations that people use to learn in the world
%% often have a partially smooth or local structure. Techniques in AI and
%% machine learning commonly use smoothness or local structure to make
%% search tractable. But the difference between this kind of learning and
%% conceptual learning is undeniable. The former is what the statistician
%% or machine learning researcher implements. The latter is what she
%% learns as a student and as a practicioner. As researchers, we try
%% different models, watch them fail or succeed, and abstract out the
%% elements of these models that capture the relevant differences. We mix
%% and match parts of different ideas, and utilize succcessful patterns
%% of reasoning that can span many problems and domains.

%% We believe that the primary characteristic of conceptual learning is
%% that it occurs over a modular and compositional computational
%% representation that supports higher-order functions. That is why we
%% chose a functional programming language as the representational
%% language of this work.

%% The experiments in this work do not contain the richness of real world
%% learning, but we believe that the learning algorithm's behavior
%% exhibits some of the characteristics we would expect of a conceptual
%% learner.

%% One characteristic of this algorithm's behavior is a rapid initial
%% increase in performance followed by a hard plateau. This plateau is
%% expected given the circular nature of the algorithm: the learner
%% inspects the region of the solution space that is typical of solutions
%% already found. We see in this phenomenon a conceptual demarcation of
%% the problem space: a coherent set of concepts or problems solving
%% techniques can define by their power the extent of problems a problem
%% solver is willing to consider. One might hope, however, that an
%% intelligent learner would recognize that such a demarcation has
%% occured and attempt to find another class of concepts for the
%% remaining unsolved tasks.

%% Another aspect of conceptual learning that we see present in this
%% algorithm's behavior is the importance of a curriculum. By this we do
%% not mean an explicit ordering of the problems in the task set, but
%% rather that a set of sufficiently simple problems must act as stepping
%% stones to the acquisition of complex representations.

>>>>

\usepackage{ijcai13}
\usepackage{times}
\usepackage{amsmath} 
\usepackage{amsfonts}
\usepackage{latexsym} 
\usepackage{ dsfont }
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{tikz}
\usepackage{tikz-qtree}
\usepackage{float}
\usepackage{circuitikz}
\usepackage{graphicx}
\usepackage[font=small]{subcaption}
\usepackage[font=small]{caption}
\usepackage{alltt}
\usepackage{dashrule}
\usepackage{array}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usetikzlibrary{arrows,automata, positioning, patterns,backgrounds}
\usetikzlibrary{arrows,shapes.gates.logic.US,shapes.gates.logic.IEC,calc}

@@

\begin{figure}
%%  \centering
  \begin{subfigure}[Before]{0.45\linewidth}
    \begin{tikzpicture}[scale=.8,
        transform shape, level/.style={sibling distance=5mm/#1, font=\scriptsize}
        ]
      \node [font=\scriptsize] (top) {$\mathbb{R}$}
      child {node [above=.3cm, left=.1cm, text width=1.8cm] (a) {$(\mathbb{R} \rightarrow \mathbb{R}) \rightarrow \mathbb{R} 
          \rightarrow \mathbb{R}$}
        child {node [left=.5cm, below=1cm, text width=3.0cm] 
          {$ \mathbf{S} : (t_2 \rightarrow t_1 \rightarrow t_0)  \rightarrow (t_2 \rightarrow t_1) 
            \rightarrow$ \\ $t_2 \rightarrow t_0$}}
        child {node [right=0cm, text width=2cm] {$ \mathbf{*}: \mathbb{R} \rightarrow \mathbb{R} \rightarrow \mathbb{R}$}}
      }
      child {node [below right=-.5cm and .2cm] (b) {$\mathbb{R} \rightarrow \mathbb{R}$}
        child {node [above=.5cm] {$ \mathbf{I}: t_0 \rightarrow t_0 \rightarrow t_0$}}
      };
    \end{tikzpicture}
    \caption{    \label{fig:clbintree}}
  \end{subfigure}
  \begin{subfigure}[After]{0.5\linewidth}
    \begin{tikzpicture}[scale=.7, transform shape, 
        level/.style={sibling distance=10mm/#1},
        every edge/.style={draw, thin}
      ]
      \node [circle,draw] (a) {}
      child {node [left=0cm] (b) {$f_1$}}
      child {node [left=.2cm] (c) {\dots} edge from parent[draw=none] }
      child {node [left=.4cm] (e) {$f_n$}}
      child {node [circle,draw, left=1cm] (d) {} edge from parent[very thick]
        child {node [circle, draw, left=.25cm] (l) {} edge from parent[draw=none]
          child {node [left=.5cm] (b) {$f_1$}} 
          child {node [left=.3cm] (c) {\dots} edge from parent[draw=none] }
          child {node [left=.3cm] (e) {$f_n$} edge from parent[very thick]}
          child {node [circle, draw, left=.25cm] (and2) {}
            child {node [circle, draw, left=.25cm](l2) {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
            child {node [circle, draw] (r2)  {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
          }
        }
        child {node [circle, draw, right=.25cm] (r) {} edge from parent[draw=none]
          child {node [circle, draw, right=.25cm] (and3) {}
            child {node [circle, draw ](l3) {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
            child {node [circle, draw, right=.2cm ] (r3)  {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
          }
          child {node [right=.3cm] (b) {$f_1$} edge from parent[very thick]} 
          child {node [right=.4cm] (c) {\dots} edge from parent[draw=none] }
          child {node [right=.6cm] (e) {$f_n$}}
        }
      };

@@

\end{subfigure}
\caption{Learning curves from Boolean function learning
  experiments. (a) Learning curves for various frontier sizes on task
  set of circuit based Boolean functions. (b) 10
  learning curves for frontier size of 2000 on a task set consisting
  of all Boolean functions of cardinality 3 or smaller. Note how most
  of the trajectories get stuck in a local minimum around
  $50\%$.\label{fig:boolLearningCurves}}
\end{figure}

@@

The problem presented here was chosen to highlight the vastness of the
search space confronting an agent with minimal background knowledge:
the reward function is binary, it is deterministic, and it cannot
provide feedback about partial solutions. An important question left
to be addressed is how to extend this algorithm to take into account
continuous rewards, noisy data, and partial solutions.

@@

At their best, human processes of discovery clearly go beyond
what our algorithm is capable of.  The algorithm can get stuck in
plateaus after a rapid initial increase in performance, due to its
circular nature: it only inspects the region of the solution space
that is typical of solutions already found.  Humans can become stuck
in similar ways, but at least sometimes they realize that they are
stuck and attempt to find another class of concepts for the remaining
unsolved tasks.  Our algorithm requires, as human learners do, a
sufficiently graded ``curriculum'', or spectrum of problems to solve --
simple problems provide necessary stepping stones to building complex
representations.  Also like humans, our algorithm can adaptively
``self-pace'' its ways through the curriculum, figuring out which
problems it is ready to tackle when, which problems are appropriately
difficult given what it currently knows.  Yet people can sometimes
more intelligently compose and modify the set of problems under
consideration, designing their own stepping stones to expand the set
of problems they can solve.  Making bootstrap learning systems smarter
in these more human-like ways is a prime target for future research.
